---
title: "US Storms and its impact on Life and Economy"
author: "Ajith Samraj"
output:
  pdf_document: default
  html_document: default
---
**Introduction**


This is the second course project in the course Reproducible Research in Coursera. Human life and Property were damaged occasionally by natural disasters such as Storms , Tsunami etc., In this analysis we focus on Storm and its impact in Life and Property.


This study is based on the analysis of the weather conditions in U.S. over a period. The data is given by National Oceanic and Atmospheric Administration (NOAA). It includes all the data such as damage to Life as well as Property.

For more details regarding the data visit                                  https://www.ncdc.noaa.gov/stormevents/


**Synopsis**


Here we are going to figure out answers for two questions

1 - which types of events cause most harm to population health?

2 - which types of events affects the economy?



**Loading Libraries**


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
```

```{r include=TRUE , cache=TRUE , results='hide'}
library(knitr)
library(ggplot2)
library(dplyr)
library(plyr)
library(stats)
library(rmarkdown)
library(tinytex)
```


**Loading Data**

```{r cache=TRUE}
#Loading the Data
RawData <- read.csv("StormData.csv.bz2")
dim(RawData)
```
 

```{r cache=TRUE}
names(RawData)
```


There are  902297 observations with 37 variables. We need only a subset of the 
whole data for analysis. 

We need only the following variables for analysing

*EVTYPE: Event Type

*FATALITIES: No of Fatalities

*INJURIES: No of Injuries

*PROGDMG: Property Damage

*PROPDMGEXP: Units for Property Damage (magnitudes - K,B,M)

*CROPDMG: Crop Damage

*CROPDMGEXP: Units for Crop Damage (magnitudes - K,BM,B)
```{r cache=TRUE}
#Selecting the Needed variables
vars <- c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
RawData <- RawData[vars]
dim(RawData)
```

**Property Damage**
```{r cache=TRUE}
unique(RawData$PROPDMGEXP)
```

Some of the value of the column PROPDMGEXP are in Letters and Some in Numbers. 
We need to Covert them to a single unit.

```{r cache=TRUE}
RawData$PROPDMGEXP <- plyr::mapvalues(RawData$PROPDMGEXP, from = c("K", "M","", "B", "m", "+", "0", "5", "6", "?", "4", "2", "3", "h", "7", "H", "-", "1", "8"), to = c(10^3, 10^6, 1, 10^9, 10^6, 0,1,10^5, 10^6, 0, 10^4, 10^2, 10^3, 10^2, 10^7, 10^2, 0, 10, 10^8))
RawData$PROPDMGEXP <- as.numeric(as.character(RawData$PROPDMGEXP))
RawData$PROPDMGTOTAL <- (RawData$PROPDMG * RawData$PROPDMGEXP)/1000000000
```


**Units of Crop Damage**
```{r cache=TRUE}
unique(RawData$CROPDMGEXP)
```

Like above some value of the column CROPDMGEXP are in Letters and Some in Numbers.
So we do conversion.

```{r cache=TRUE}
RawData$CROPDMGEXP <- mapvalues(RawData$CROPDMGEXP, from = c("","M", "K", "m", "B", "?", "0", "k","2"),
                     to = c(1,10^6, 10^3, 10^6, 10^9, 0, 1, 10^3, 10^2))
RawData$CROPDMGEXP <- as.numeric(as.character(RawData$CROPDMGEXP))
RawData$CROPDMGTOTAL <- (RawData$CROPDMG * RawData$CROPDMGEXP)/1000000000
```

# Events with more Fatality rate
**Fatality rate**

Analysing which events cause most Fatalities . FATILITIES is the factor variable. Almost 985 events were 
recorded by NOAA.

```{r cache=TRUE}
sumFatalities <- aggregate(FATALITIES ~ EVTYPE, data = RawData,  FUN="sum")
dim(sumFatalities)
```

Looking top 5 fatal events.
```{r cache=TRUE}
fatalities5events <- sumFatalities[order(-sumFatalities$FATALITIES), ][1:5, ]
dim(fatalities5events)
fatalities5events
```

**Fatality Plot**

Plot of the Top 5 fatal events.
 
```{r cache=TRUE}
par(mar=c(12, 3, 3, 1))
barplot(fatalities5events$FATALITIES, names.arg = fatalities5events$EVTYPE, las = 3,
        main = "Top 5 Fatalities", ylab = "No.of Fatalities")
```


**Injury rate**

Events which are causing injuries to human life.
```{r cache=TRUE}
sumInjuries <- aggregate(INJURIES ~ EVTYPE, data = RawData,  FUN="sum")
dim(sumInjuries)
```

Sorting Top 5 injury causing events.

```{r cache=TRUE}
injuries5events <- sumInjuries[order(-sumInjuries$INJURIES), ][1:5, ]
dim(injuries5events)
```

Top 5 Injury events

```{r cache=TRUE}
injuries5events
```

Plot of the top 5 injury causing events

```{r cache=TRUE}
par(mar=c(10,4,4,4))
barplot(injuries5events$INJURIES, names.arg = injuries5events$EVTYPE, las = 3, 
        main = "Top 5 Injuries Events", ylab = "No.of Injuries")
```

# Events with more Economic damage

We have to rely on PROPDMG (Property Damage) and CROPDMG (Crop Damage) to figue out the Economic impacts.

**Crop Damage**

Crop damage is given by the variable CROPDMG.

```{r cache=TRUE}
sumCropDamage <- aggregate(CROPDMGTOTAL ~ EVTYPE, data = RawData,  FUN="sum")
dim(sumCropDamage)
```

Top 5 Crop damaging events
```{r cache=TRUE}
cropdmg5Total <- sumCropDamage[order(-sumCropDamage$CROPDMGTOTAL), ][1:5, ]
cropdmg5Total
```

Plot of the Top 5 Crop damaging events

```{r cache=TRUE}
par(mar=c(7,4,4,4))
barplot(cropdmg5Total$CROPDMGTOTAL, names.arg = cropdmg5Total$EVTYPE, las = 2, 
        main = "Top 5 Crop Damaging Events",
        ylab = "Damage Crop Values (in Billions)")
```


**Property Damage Events**

```{r cache=TRUE}
sumPropertyDamage <- aggregate(PROPDMGTOTAL ~ EVTYPE, data = RawData,  FUN="sum")
dim(sumPropertyDamage)
```

Top 5 Property damage events

```{r cache=TRUE}
propdmg5Total <- sumPropertyDamage[order(-sumPropertyDamage$PROPDMGTOTAL), ][1:5, ]
propdmg5Total
```

Plot for Top 5 property Damaging events

```{r cache=TRUE}
par(mar=c(12,4,4,4))
barplot(propdmg5Total$PROPDMGTOTAL, names.arg = propdmg5Total$EVTYPE, las = 3, 
        main = "Top 5 Property damaging events",
        ylab = "Damage Property Values (in Billions)")
```

# Results

**Question - 1 **


Which Event causes more Fatalities ?


The results tells us that **Tornados** causes the highest number of Fatalities and Injuries.


**Question - 2 **


Which Event causes more Economic consequences ?


Since we analaysed on two perspectives we got two results according to the property.

      
The results tells us that **Floods** causes highest **Property Damage**.


The results tells us that **Droughts** causes highest **Crop damages**.
      
